---
title: "A3: Incarceration"
author: "Thao-Nguyen Ho"
date: "8/6/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

### Introduction

Incarceration is a big problem in the U.S. and negatively impacts communities,
especially minorities. Understanding this issue can help design better policies
to reduce incarceration rates across the country and break down the racial and
gender disparities embedded in this issue. To explore patterns of incarceration
in the U.S., I’ll be looking at this [dataset about jail population per county/state](https://github.com/melaniewalsh/Neat-Datasets/blob/main/us-jail-pop.csv?raw=true). Due to having a lot of missing values in the years before 1990, I’ll only
focus on data that are recorded from 1990 to 2018, specifically for the
population aged 15 to 64 as this group is the most likely to be incarcerated.

The measure of incarceration I’ll be focusing on is the incarceration percentage
of a particular group population. One of the variables I’ll be analyzing is race
because I’m interested in seeing how racial disparities have changed over time
and which groups have higher or increasing rates of incarceration. Another
variable I’ll be looking at is gender, specifically the relationship between 
males' and females' jail incarceration rate, to better understand how gender
differences and biases can affect the incarceration rate. Lastly, I’ll be
examining the incarceration of the most affected group across different states
of the U.S. to discover any geographical clusters and factors that can 
contribute to the inequalities in the criminal justice system.

### Summary Information

```{r, echo=FALSE, message=FALSE}
source("summary.R")
```

In the most recent year recorded, 2018, there was an average of `r avg_pop`
people per state that are in jail in the U.S., with the median at `r median_pop`
people. From 1990 to 2018, the group with the highest incarceration rate is
Black, such that on average, `r max_rate`% of the Black population is in jail.
While AAPI people has the lowest incarceration rate of `r min_rate`%. My
analysis also showed that approximately `r male_mean_pct`% of male population
are in jail each year. Compared to `r female_mean_pct`% of female population
that are in jail, males are `r gender_ratio` times more likely to get
incarcerated. Further analyzing shows that the state with the highest percentage
of its Black population in jail is `r max_pct_state`. Due to missing data for
some states, some states has 0% of its Black population in jail but this is
inaccurate and need further analysis from other datasets.

### The Dataset

Who collected the data?

- The data was collected by the [Vera Institute](https://github.com/vera-institute/incarceration-trends)

How was the data collected or generated?

- From the [Codebook](https://github.com/vera-institute/incarceration-trends/blob/master/incarceration_trends-Codebook.pdf), the data was collected from datasets provided by
the U.S. Department of Justice Bureau of Justice Statistics (BJS), the Deaths
in Custody Reporting Program (DCRP), and the Annual Survey of Jails.

Why was the data collected?

- The data was collected to document and inform the public of mass incarceration
in the U.S. It can be used by others as a tool to understand who and how many
are in jails/prisons and to analyze patterns that can reveal the disparities in
the criminal justice system.

How many observations (rows) are in your data?

- There are 153811 observations (rows) in the dataset.

How many features (columns) are in the data?

- There are 28 features (columns) in the dataset.

What, if any, ethical questions or questions of power do you need to consider when working with this data?

- When working with this data, some ethical questions I need to consider are
related to the privacy of the individuals. These questions include what kinds of
information are collected from each individual and how this information is
stored, and whether the participants are aware of this whole process. Because
the values presented in this data are mostly numbers, how can we make sure that
they are not over or under estimate for certain gender or racial groups, and
whether there’re biases presented in the data. 

What are possible limitations or problems with this data?   (at least 200 words)

- A possible problem with this dataset is its reliability and usefulness because
of how many missing values are in the data. This can skew our results and lead
to inaccurate analysis. Missing values can also lead to a group being
underrepresented in the data. For example from the summary given above, because
there were no incarceration data recorded for some states, they seem to be
having the lowest incarcerated population, but this interpretation is
inaccurate. To avoid working with too many missing values, we can try to collect
those values ourselves or focus only on a subset or section of data that has
sufficient values to work with. A possible limitation with this data is how much
details it can provide because it might not have sufficient features to draw
meaningful correlations from. Because factors like the reasons people got
incarcerated can have a big influence on the increasing or decreasing of
incarceration rate for certain gender and racial groups. But this can be hard to
collect and store on a large scale. 

### Trends Over Time Chart

```{r line_chart, echo = FALSE, message = FALSE}
source("chart1.R")
plot(line_chart)
```

This chart visualizes the jail incarceration percentage of different racial
groups from the age of 15 to 64 in the U.S. 1990-2018. A racial group’s
percentage represents the percentage of people in jail in relation to that
group's total population.

#### Purpose

By comparing the jail incarceration percentage over time of different racial
groups, this chart seeks to express the trends of the incarcerated population
and which group has the highest incarceration rates. The chart allows us to see
which year there was a sudden spike in the incarceration rate and allows for
further analysis of the racial disparities in the criminal justice system.
Understanding these patterns can provide insights into creating programs that
help lower the incarceration rates of the most affected groups.

#### Patterns

- From 1990 to 2018, **Black** people have the **highest** jail incarceration
percentage. The group reached the highest rate at around **2006**, during
which **1.15%** of its population are in jail. While **AAPI** people have the
**lowest** jail incarceration percentage.
  - White people have the second lowest trend and before the year 2000, Latinx
  people were the second highest followed by Native people but their spot got
  switched from 2000 and onward.
- **Black, Latinx, and AAPI** groups all have a **decreasing** trend in the
percentage of their people in jail after 2008. While we can see an 
**increasing** trend for **Native and White** groups.
  - This might be the result of different reforms happening during that period
  and an increase in programs for reducing crimes. But more focus needs to be
  put on Native people to help reduce or keep their jail incarceration
  percentage from rising higher. 


### Variable Comparison Chart

```{r scatter_chart, echo = FALSE, message = FALSE}
source("chart2.R")
plot(scatter_chart)
```

This chart visualizes the relationship between males' and females' jail
incarceration percentages from the age of 15 to 64 in the U.S. 1990-2018. A
gender’s percentage represents the percentage of people in jail in relation to
that gender’s total population. 

#### Purpose

By showing the correlation between males' and females' incarceration
percentages, this chart seeks to show how these two variables are related to one
another and the direction that they both are going over the years. It allows us
to see which gender is increasing or decreasing at a faster rate and whether
this affect the outcome of the other gender. Understanding these patterns can
help make changes to policies that better fit the needs of both groups and
explore gender biases in the criminal justice system.

#### Patterns

- At the **highest** incarceration percentage, approximately **0.672% of total**
**male population** are in jail while only **0.098% of females**. Males are
**7 times** more likely than females to be incarcerated.
- There is a **positive** correlation between males' and females' jail
incarceration rates except for some outliers in the most recent years.
  - In the most recent year, female incarceration percentage seems to stay the
  same at its highest while male incarceration percentage had a slight drop from
  its highest to around 0.58%.
  - This might indicate that policy changes to help bring down incarceration
  rates didn’t take females into account, which may be because they don’t
  represent the majority of jail population.

### Map

```{r pop_map, echo = FALSE, message = FALSE}
source("map.R")
plot(pop_map)
```

This chart visualizes the jail incarceration percentage of the Black population
from the age of 15 to 64 in 2018 across different states of the U.S. 

#### Purpose

By showing how the Black jail incarceration percentage varies geographically,
this chart seeks to show which state has the highest percentage of its Black
population incarcerated and which one has the lowest. It allows us to see
whether there’re patterns across the different states, like geographical
clusters of Black incarceration, and to question and explore why such patterns
exist.

#### Patterns

- Some states like **Connecticut, Delaware, and Vermont** have **0%** of their
Black population incarcerated, indicating that there are missing data for these
states. 
  - Alaska and Hawaii are missing from the map due to limited technical skills.
- **Wyoming** is the state with the **highest** Black jail incarceration
percentage, in which **2.43%** of its Black population from the age of 15 to 64
are in jail in 2018.
  - The states around Wyoming also have a high percentage. More context about
  each state needs to be explored to understand why these states have high Black
  incarceration percentages.

