---
title: "A3: Incarceration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

### Introduction

An introduction of the problem domain and a description of the variable(s) you are choosing to analyze (and why!)

Variables: race, gender, region

### Summary Information

```{r, echo=FALSE, message=FALSE}
source("summary.R")
```

### The Dataset

Who collected the data?  
How was the data collected or generated?  
Why was the data collected?  
How many observations (rows) are in your data?  
How many features (columns) are in the data?  
What, if any, ethical questions or questions of power do you need to consider when working with this data?  
What are possible limitations or problems with this data?   (at least 200 words)


### Trends Over Time Chart

```{r line_chart, echo = FALSE, message = FALSE}
source("chart1.R")
plot(line_chart)
```

This chart visualizes...

#### Purpose

#### Patterns

### Variable Comparison Chart

```{r scatter_chart, echo = FALSE, message = FALSE}
source("chart2.R")
plot(scatter_chart)
```

This chart visualizes...

#### Purpose

#### Patterns

### Map

```{r pop_map, echo = FALSE, message = FALSE}
source("map.R")
plot(pop_map)
```

This chart visualizes...

#### Purpose

#### Patterns

