---
title: "A3: Incarceration"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE)
```

### Introduction

An introduction of the problem domain and a description of the variable(s) you are choosing to analyze (and why!)

Variables: race, gender, region

### Summary Information

```{r, echo=FALSE, message=FALSE}
source("summary.R")
```

### The Dataset

Who collected the data?  
How was the data collected or generated?  
Why was the data collected?  
How many observations (rows) are in your data?  
How many features (columns) are in the data?  
What, if any, ethical questions or questions of power do you need to consider when working with this data?  
What are possible limitations or problems with this data?   (at least 200 words)


### Trends Over Time Chart

```{r line_chart, echo = FALSE, message = FALSE}
source("chart1.R")
plot(line_chart)
```

This chart visualizes the jail incarceration percentage of different racial
groups from the age of 15 to 64 in the U.S. 1990-2018. A racial group’s
percentage represents the percentage of people in jail in relation to that
group's total population.

#### Purpose

By comparing the jail incarceration percentage over time of different racial
groups, this chart seeks to express the trends of the incarcerated population
and which group has the highest incarceration rates. The chart allows us to see
which year there was a sudden spike in the incarceration rate and allows for
further analysis of the racial disparities in the criminal justice system.
Understanding these patterns can provide insights into creating programs that
help lower the incarceration rates of the most affected groups.

#### Patterns

- From 1990 to 2018, **Black** people have the **highest** jail incarceration
percentage. The group reached the highest rate at around **2006**, during
which **1.15%** of its population are in jail. While **AAPI** people have the
**lowest** jail incarceration percentage.
  - White people have the second lowest trend and before the year 2000, Latinx
  people were the second highest followed by Native people but their spot got
  switched from 2000 and onward.
- **Black, Latinx, and AAPI** groups all have a **decreasing** trend in the
percentage of their people in jail after 2008. While we can see an 
**increasing** trend for **Native and White** groups.
  - This might be the result of different reforms happening during that period
  and an increase in programs for reducing crimes. But more focus needs to be
  put on Native people to help reduce or keep their jail incarceration
  percentage from rising higher. 


### Variable Comparison Chart

```{r scatter_chart, echo = FALSE, message = FALSE}
source("chart2.R")
plot(scatter_chart)
```

This chart visualizes the relationship between males' and females' jail
incarceration percentages from the age of 15 to 64 in the U.S. 1990-2018. A
gender’s percentage represents the percentage of people in jail in relation to
that gender’s total population. 

#### Purpose

By showing the correlation between males' and females' incarceration
percentages, this chart seeks to show how these two variables are related to one
another and the direction that they both are going over the years. It allows us
to see which gender is increasing or decreasing at a faster rate and whether
this affect the outcome of the other gender. Understanding these patterns can
help make changes to policies that better fit the needs of both groups and
explore gender biases in the criminal justice system.

#### Patterns

- At the **highest** incarceration percentage, approximately **0.672% of total**
**male population** are in jail while only **0.098% of females**. Males are
**7 times** more likely than females to be incarcerated.
- There is a **positive** correlation between males' and females' jail
incarceration rates except for some outliers in the most recent years.
  - In the most recent year, female incarceration percentage seems to stay the
  same at its highest while male incarceration percentage had a slight drop from
  its highest to around 0.58%.
  - This might indicate that policy changes to help bring down incarceration
  rates didn’t take females into account, which may be because they don’t
  represent the majority of jail population.

### Map

```{r pop_map, echo = FALSE, message = FALSE}
source("map.R")
plot(pop_map)
```

This chart visualizes the jail incarceration percentage of the Black population
from the age of 15 to 64 in 2018 across different states of the U.S. 

#### Purpose

By showing how the Black jail incarceration percentage varies geographically,
this chart seeks to show which state has the highest percentage of its Black
population incarcerated and which one has the lowest. It allows us to see
whether there’re patterns across the different states, like geographical
clusters of Black incarceration, and to question and explore why such patterns
exist.

#### Patterns

- Some states like **Connecticut, Delaware, and Vermont** have **0%** of their
Black population incarcerated, indicating that there are missing data for these
states. 
  - Alaska and Hawaii are missing from the map due to limited technical skills.
- **Wyoming** is the state with the **highest** Black jail incarceration
percentage, in which **2.43%** of its Black population from the age of 15 to 64
are in jail in 2018.
  - The states around Wyoming also have a high percentage. More context about
  each state needs to be explored to understand why these states have high Black
  incarceration percentages.

